---
title: "chapter 3"
author: "Imani Russell"
date: "4/21/2021"
output:
  pdf_document: default
  html_document: default
---

########## skip all of this code ######################################
Read in (and comment out) 2 R scripts that are called - I can't figure out how else to get the initial setup to work and include all functions needed for chapter 3


```{r Standard Graphical Pars.R}
# set_graph_pars <- function(ptype = "panel4")
# {
#     mgp <- c(2.5, 1, 0); mar <- c(4, 4, 1.5, 1); oma <- c(0, 0, 0, 0)
#     switch(ptype,
#            panel4 = par(mfrow=c(2,2), mgp=mgp, mar=mar, pty="s",
#                         oma=oma, bty="L", cex.lab =1.2),
#            panel3 = par(mfrow=c(1,3), mgp=mgp, mar=mar, pty="s",
#                         oma=oma, bty="L", cex.axis=0.65),
#            panel2 = par(mfrow=c(1,2), mgp=mgp, mar=mar, pty="s",
#                         oma=oma, bty="L", cex.axis=0.85),
#            panel1 = par(mfrow=c(1,1), mgp=mgp, mar=mar, pty="s",
#                         oma=oma, bty="L", cex.axis=0.85))
# }
# 
# add_panel_label <- function(ltype="a")
# {
#     text <- paste(LETTERS[letters==ltype], ")", sep="")
#     mtext(text=text, side=3, adj=0)
# }
# 
# ## change to 'TRUE' to look at the formatting...
# if (FALSE) {
# 
#     plot_dummy <- function()
#         plot(1:100, 1:100, type="n", xlab="x variable", ylab="y variable")
#     ## 4 panels
#     dev.new(width=8, height=7.5)
#     set_graph_pars(ptype = "panel4")
#     plot_dummy(); add_panel_label("a")
#     plot_dummy(); add_panel_label("b")
#     plot_dummy(); add_panel_label("c")
#     plot_dummy(); add_panel_label("d")
#     ## 2 panels
#     dev.new(width=8, height=4)
#     set_graph_pars(ptype = "panel2")
#     plot_dummy(); add_panel_label("a")
#     plot_dummy(); add_panel_label("b")
#     ## 2 panels
#     dev.new(width=8, height=4)
#     set_graph_pars(ptype = "panel1")
#     plot_dummy()
#     ## clean up
#     rm(plot_dummy)
# }

```

```{r Monocarp Simulate IBM.R}
## ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
## Section 1 - Simulate the IPM
## ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# library(doBy)
# set.seed(53241986)
# 
# # Set simulation parameters
# init.pop.size <- 250
# n.yrs <- 100
# 
# # initial population sizes and ages
# z <- rnorm(init.pop.size, mean = m.par.true["rcsz.int"], sd = m.par.true["rcsz.sd"])
# age <- rep(0, init.pop.size)
# 
# ## calculate initial pop size and mean size
# 
# pop.size.t <- init.pop.size
# mean.z.t <- mean(z)
# mean.age.t <- mean(age)
# 
# ## iterate the model using the 'true' parameters and store data in a
# ## data.frame
# yr <- 1
# while (yr != n.yrs & length(z) < 1500000) {
#   
#   ## calculate population size
#   pop.size <- length(z)
#   
#   ## generate binomial random number for the probability of flowering,
#   ## where the probability of flowering depends on your size z, this is a
#   ## vector of 0's and 1's, you get a 1 if you flower
#   Repr <- rbinom(n = pop.size, prob = p_bz(z, m.par.true), size = 1)
#   
#   ## number of plants that flowered
#   num.Repr <- sum(Repr)
#   
#   ## calculate seed production
#   Seeds <- rep(NA, pop.size)
#   
#   ## we'll assume plant make a Poisson distributed number of seeds with a
#   ## mean given by exp(params['seed.int']+params['seed.size'] * z) rpois
#   ## generated Poisson distributed random numbers
#   Seeds[Repr == 1] <- rpois(num.Repr, b_z(z[Repr == 1], m.par.true))
#   
#   ## generate the number of recruits
#   Recr <- ifelse(num.Repr == 0, 0, rbinom(1, sum(Seeds, na.rm = TRUE), 
#                                           m.par.true["p.r"]))
#   
#   ## generate new recruit sizes rnorm generated normally distributed
#   ## random numbers
#   Rcsz <- rnorm(Recr, mean = m.par.true["rcsz.int"], sd = m.par.true["rcsz.sd"])
#   
#   ## for the non-reproductive plants generate random number for survival
#   Surv <- rep(NA, pop.size)
#   Surv[Repr == 0] <- rbinom(n = pop.size - num.Repr, 
#                             prob = s_z(z[Repr == 0], m.par.true), size = 1)
#   num.die <- sum(Surv == 0, na.rm = TRUE)
#   
#   ## index for individuals that did not flower and survived
#   i.subset <- which(Repr == 0 & Surv == 1)
#   
#   ## let them grow
#   z1 <- rep(NA, pop.size)
#   E.z1 <- m.par.true["grow.int"] + m.par.true["grow.z"] * z[i.subset]
#   z1[i.subset] <- rnorm(n = pop.size - num.Repr - num.die, mean = E.z1, 
#                         sd = m.par.true["grow.sd"])
#   alive <- (Repr == 0 & Surv == 1)
#   
#   ## store the simulation data, we'll use this later
#   
#   if (yr == 1) {
#     sim.data <- data.frame(z = z, Repr, Seeds, Surv, z1 = z1, age = age, 
#                            alive = alive, yr = 1)
#   } else {
#     sim.data <- rbind(sim.data, data.frame(z = z, Repr, Seeds, Surv, 
#                                            z1 = z1, age = age, alive = alive, yr = yr))
#   }
#   
#   ## create new population size vector
#   z1 <- c(Rcsz, z1[i.subset])
#   age1 <- c(rep(0, Recr), age[i.subset] + 1)
#   
#   pop.size.t <- c(pop.size.t, length(z1))
#   mean.z.death.t <- if (yr == 1) 
#     mean(z[!(Repr == 0 & Surv == 1)]) else c(mean.z.death.t, mean(z[!(Repr == 0 & Surv == 1)]))
#   mean.age.death.t <- if (yr == 1) 
#     mean(age[!(Repr == 0 & Surv == 1)]) else c(mean.age.death.t, mean(age[!(Repr == 0 & Surv == 1)]))
#   mean.z.t <- if (yr == 1) 
#     mean(z) else c(mean.z.t, mean(z))
#   mean.fl.z.t <- if (yr == 1) 
#     mean(z[Repr == 1]) else c(mean.fl.z.t, mean(z[Repr == 1]))
#   mean.fl.age.t <- if (yr == 1) 
#     mean(age[Repr == 1]) else c(mean.fl.age.t, mean(age[Repr == 1]))
#   
#   
#   z <- z1
#   age <- age1
#   
#   cat(paste(yr, mean.age.death.t[yr], mean.fl.age.t[yr], "\n", sep = " "))
#   
#   yr <- yr + 1
#   
# }




```

```{r Monocarp Demog Funs.R}
## IBM to generate the data for the simple recruitment limited IPM - there's no recruitment limitation 1st
## Size is on a log scale so don't worry about the -ive ones!

## The parameters and life cycle follow Kachi and Hirose J.Ecol. 1985, see Rees and Rose PRSB 2002, but for
## simplicity we have used a logistic function for survival rather than the linear function with an upper bound
## as in the previous papers

## ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
## Section 1 - Define the demographic functions and parameters
## ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

## Define the true parameter vector, each parameter is given a name so the formulae, below, are easier to
## read. We'll use 'model.term' scheme to name elements of the vector

# m.par.true <- c(## survival
#                 surv.int  =  -0.65,
#                 surv.z    =   0.75,
#                 ## flowering
#                 flow.int  = -18.00,
#                 flow.z    =   6.9,
#                 ## growth
#                 grow.int  =   0.96,
#                 grow.z    =   0.59,
#                 grow.sd   =   0.67,
#                 ## recruit size
#                 rcsz.int  =   -.08, 
#                 rcsz.sd   =   0.76,
#                 ## seed size
#                 seed.int  =   1.00,
#                 seed.z    =   2.20,
#                 ## recruitment probability
#                 p.r       =   0.007)  
# 
# 
# ##
# ## Define the various demographic functions, we pass a vector of parameters "m.par" to each function
# ## this makes it easier to compare the results of different paramter sets, say the true values and
# ## estimated ones
# ##
# 
# ## Growth function, given you are size z now returns the pdf of size z1 next time
# 
# G_z1z <- function(z1, z, m.par){
#     mu <- m.par["grow.int"] + m.par["grow.z"] * z   # mean size next year
#     sig <- m.par["grow.sd"]      # sd about mean
#     p.den.grow <- dnorm(z1, mean = mu, sd = sig) # pdf that you are size z1 given you were size z
#     return(p.den.grow)
# }
# 
# ## Survival function, logistic regression
# 
# s_z <- function(z, m.par){
#     linear.p <- m.par["surv.int"] + m.par["surv.z"] * z  # linear predictor
#     p <- 1/(1+exp(-linear.p))     # logistic transformation to probability
#     return(p)
# }
# 
# ## Probability of flowering function, logistic regression
# 
# p_bz <- function(z, m.par){
#     linear.p <- m.par["flow.int"] + m.par["flow.z"] * z   # linear predictor
#     p <- 1/(1+exp(-linear.p))  # logistic transformation to probability
#     return(p)
# }
# 
# ## Seed production function
# 
# b_z <- function(z, m.par)
# {
#     N <- exp(m.par["seed.int"] + m.par["seed.z"] * z) # seed production of a size z plant
#     return(N)
# }
# 
# ## Recruit size pdf
# 
# c_0z1 <- function(z1, m.par)
# {
#     mu <- m.par["rcsz.int"]
#     sig <- m.par["rcsz.sd"]
#     p.deRecr <- dnorm(z1, mean = mu, sd = sig)    # pdf of a size z1 recruit
#     return(p.deRecr)
# }
# 
# #### fit models capturing size-dependent rates of survival and reproduction
# #(** didn't we just do this??**)
# 
# sim.data.noRepr <- subset (sim.data, Repr == 0)
# 
# ## ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# ## Section 2 - Functions to build IPM kernels P, F, and K
# ## ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# 
# ## Define the survival kernel
# P_z1z <- function (z1, z, m.par) {
# 
#     return((1 - p_bz(z, m.par)) * s_z(z, m.par) * G_z1z(z1, z, m.par))
# 
# }
# 
# ## Define the fecundity kernel
# F_z1z <- function (z1, z, m.par) {
# 
#     return( p_bz(z, m.par) * b_z(z, m.par) * m.par["p.r"] * c_0z1(z1, m.par))
# 
# }
# 
# mk_K <- function(m, m.par, L, U) {
# 
# 	# mesh points 
# 	h <- (U - L)/m
# 	meshpts <- L + ((1:m) - 1/2) * h
# 	P <- h * (outer(meshpts, meshpts, P_z1z, m.par = m.par))
# 	F <- h * (outer(meshpts, meshpts, F_z1z, m.par = m.par))
# 	K <- P + F
# 	return(list(K = K, meshpts = meshpts, P = P, F = F))
# }
# 
# 
# # flexible size limits, defaults set for Oenothera model
# mk_K_ceiling <- function(m, m.par, L, U, U1 = U) {
# 	# mesh points 
# 	h <- (U - L)/m;
# 	meshpts <- L + ((1:m) - 1/2) * h;
# 	P <- h * (outer(meshpts, pmin(meshpts,U1), P_z1z, m.par = m.par));
# 	F <- h * (outer(meshpts, pmin(meshpts,U1), F_z1z, m.par = m.par));
# 	K <- P + F;
# 	return(list(K = K, meshpts = meshpts, P = P, F = F))
# }
# 
# #Function to calculate mean and variance in reproductive output for a size z
# 
# get_mean_var_Repr <- function(init.z,n.samp) {
# 
# # initial population sizes and ages
# z   <- rep(init.z,1000)
# Repr.out <- NULL
# 
# repeat {
# 
#     ## calculate population size
#     pop.size <- length(z)
# 
#     ## generate binomial random number for the probability of flowering, where the probability of flowering
#     ## depends on your size z, this is a vector of 0's and 1's, you get a 1 if you flower
#     Repr <- rbinom(n=pop.size, prob=p_bz(z, m.par.true), size=1)
# 
#     ## number of plants that flowered
#     num.Repr <- sum(Repr)
# 
#     if(num.Repr>0) {
#     	Seeds <- rpois(num.Repr, m.par.true["p.r"] * b_z(z[Repr==1],m.par.true))
#     	Repr.out <- c(Repr.out,Seeds)
#     }
# 
#     ## generate new recruit sizes
#     ## rnorm generated normally distributed random numbers
#     Rcsz <- rep(init.z,100)
# 
#     ## for the non-reproductive plants generate random number for survival
#     Surv <- rep(NA, pop.size)
#     Surv[Repr==0] <- rbinom(n = pop.size - num.Repr, prob = s_z(z[Repr==0], m.par.true), size = 1)
#     num.die <- sum(Surv==0, na.rm=TRUE)
#     
#     if(num.die>0) Repr.out <- c(Repr.out,rep(0,num.die))
# 
#     ## index for individuals that did not flower and survived
#     i.subset <- which(Repr==0 & Surv==1)
# 
#     ## let them grow
#     E.z1 <- m.par.true["grow.int"]+m.par.true["grow.z"]*z[i.subset]
#     z1 <- rnorm(n = pop.size - num.Repr - num.die, mean = E.z1, sd = m.par.true["grow.sd"])
#     
#     z <- c(Rcsz, z1)
# 	 
#     if(length(Repr.out)>n.samp) break
# 
# }
# 
# return(c(mean(Repr.out),var(Repr.out),mean(Repr.out>0),var(Repr.out>0)))
# 
# }
```


### From Chapter 2, Cherie's adaptation of the code:
### we'll need this to run the code in Ch 3

Do not include this in the knitted final version! 
```{r, echo=FALSE, results=FALSE}
rm(list = ls(all = TRUE))

library(doBy)
set.seed(53241986)

# Set simulation parameters
init.pop.size <- 250
n.yrs <- 100

source("~/Dropbox/dissertation research/IPM/ipm_book/Rcode/utilities/Standard Graphical Pars.R")

source("~/Dropbox/dissertation research/IPM/ipm_book/Rcode/c2/Monocarp Simulate IBM.R")
source("~/Dropbox/dissertation research/IPM/ipm_book/Rcode/c2/Monocarp Demog Funs.R")
cat(pop.size.t, "\n") #read population sizes with space in between?


# trim an initial transient off the simulation
sim.data <- sim.data[sim.data$yr > 10, ]
sim.data$yr <- sim.data$yr - 10
head(sim.data)

# Extract 1000 observations to use as our data set
sample.index <- sample(1:nrow(sim.data), size = 1000, replace = FALSE)
sim.data <- sim.data[sample.index, ]
dim(sim.data)
```

```{r, echo=FALSE, results=FALSE, Fit statistical models to simulated data}
# Use simulated data to construct an IPM, assuming we know the order of events first a bit of house keeping for plotting - chop up the sizes into classes and sort the data by size

sim.data <- transform(sim.data, z.classes = cut(z, 20))
sim.data <- sim.data[order(sim.data$z), ]
head(sim.data)

# 2.5.3 Demographic Analyses using lm and glm
## fit the functions and plot some graphs

## 1 - growth size
mod.Grow <- lm(z1 ~ z, data = sim.data)
summary(mod.Grow)

# growth plot
plot(z1 ~ z, data = sim.data, pch = 16, cex = 0.25, 
     xlab = expression("Size t, " * italic(z)), 
     ylab = expression("Size t+1, " * italic(z) * "'"), 
     main="Growth")
abline(mod.Grow, col = "red")
lines(z,z,col="gray")

## 2 - probability of flowering
mod.Repr <- glm(Repr ~ z, family = binomial, data = sim.data)
summary(mod.Repr)

# flowering plot
Repr.ps <- summaryBy(z + Repr ~ z.classes, data = sim.data)
plot(Repr ~ z, data = sim.data, cex = 0.7, 
     xlab = expression("Size t, " *italic(z)), 
     ylab = "Probability of flowering", col="gray")
points(Repr.mean ~ z.mean, pch = 19, col="black", data = Repr.ps)
lines(fitted(mod.Repr) ~ z, data = sim.data, col = "red")

## 3 - probability of survival
sim.data.noRepr <- subset(sim.data, Repr == 0) # select non-reproductive flowers - flowering fatal and occurs before all other events in life cycle
mod.Surv <- glm(Surv ~ z, family = binomial, data = sim.data.noRepr)
summary(mod.Surv)

# survival plot
surv.ps <- summaryBy(z + Surv ~ z.classes, data = sim.data.noRepr, na.rm = TRUE)

plot(Surv ~ z, data = sim.data.noRepr, col="gray", cex = 0.7,
     xlab = expression("Size t, " *italic(z)), 
     ylab = "Probability of survival")
points(Surv.mean ~ z.mean, data = surv.ps, pch = 19)

lines(fitted(mod.Surv) ~ z, data = sim.data.noRepr, col = "red")

## 4 - seed production
sim.data.Repr <- subset(sim.data, Repr == 1)
mod.Seeds <- glm(Seeds ~ z, family = poisson, data = sim.data.Repr)
summary(mod.Seeds)

# seed production plot
plot(log(Seeds) ~ z, data = sim.data.Repr, pch = 19, 
     xlab = expression("Size t, " *italic(z)), 
     ylab = "Seeds production")

abline(mod.Seeds, col = "red")

## 5 - recruit size
sim.data.Rec <- subset(sim.data, age == 0)
mod.Rcsz <- lm(z ~ 1, sim.data.Rec)
summary(mod.Rcsz)

# recruit size plot
hist(sim.data.Rec$z, xlab="Size of Recruits")

## 6 - establishment probability 
p.r.est <- as.numeric(Recr)/sum(Seeds, na.rm = TRUE)
p.r.est

#store the estimated parameters
m.par.est <- c(surv = coef(mod.Surv), flow = coef(mod.Repr), grow = coef(mod.Grow), grow.sd = summary(mod.Grow)$sigma, rcsz = coef(mod.Rcsz), rcsz.sd = summary(mod.Rcsz)$sigma, seed = coef(mod.Seeds), p.r = p.r.est)

m.par.est

# This is just giving the parameters the same names as the true parameters
names(m.par.est) <- names(m.par.true)
m.par.est
```


```{r, echo=FALSE, results=FALSE}
# Construct Kernels and projection population size
# this is just finding the min and max of z in sim.data
nBigMatrix <- 250
min.size <- with(sim.data, min(z))
min.size
max.size <- with(sim.data, max(z))
max.size
# so let's set the lower and upper limits for the size range at -2.65, and 4.5 so slightly smaller/bigger than observed.
```


```{r, echo=FALSE, results=FALSE}
# 2.5.4 Implementing the IPM
#From Monocarp Demog Funs.R
#Functions to build IPM kernels P, F, and K

## Define the survival kernel
P_z1z <- function (z1, z, m.par) {

    return((1 - p_bz(z, m.par)) * s_z(z, m.par) * G_z1z(z1, z, m.par))

}

## Define the fecundity kernel
F_z1z <- function (z1, z, m.par) {

    return( p_bz(z, m.par) * b_z(z, m.par) * m.par["p.r"] * c_0z1(z1, m.par))

}

## Make Kernel
mk_K <- function(m, m.par, L, U) {

	# mesh points 
	h <- (U - L)/m
	meshpts <- L + ((1:m) - 1/2) * h
	P <- h * (outer(meshpts, meshpts, P_z1z, m.par = m.par))
	F <- h * (outer(meshpts, meshpts, F_z1z, m.par = m.par))
	K <- P + F
	return(list(K = K, meshpts = meshpts, P = P, F = F))
}
```

```{r, echo=FALSE, results=FALSE}
## Define the various demographic functions.
###We pass a vector of parameters "m.par" to each function. This makes it easier to compare the results of different paramter sets, say the true values and estimated ones

## Growth function, given you are size z now returns the pdf of size z1 next time
G_z1z <- function(z1, z, m.par)
{
    mu <- m.par["grow.int"] + m.par["grow.z"] * z           # mean size next year
    sig <- m.par["grow.sd"]                                    # sd about mean
    p.den.grow <- dnorm(z1, mean = mu, sd = sig)             # pdf that you are size z1 given you were size z
    return(p.den.grow)
}

## Survival function, logistic regression
s_z <- function(z, m.par)
{
    linear.p <- m.par["surv.int"] + m.par["surv.z"] * z  # linear predictor
    p <- 1/(1+exp(-linear.p))                            # logistic transformation to probability
    return(p)
}

## Probability of flowering function, logistic regression
p_bz <- function(z, m.par)
{
    linear.p <- m.par["flow.int"] + m.par["flow.z"] * z      # linear predictor
    p <- 1/(1+exp(-linear.p))                                # logistic transformation to probability
    return(p)
}

## Seed production function
b_z <- function(z, m.par)
{
    N <- exp(m.par["seed.int"] + m.par["seed.z"] * z)    # seed production of a size z plant
    return(N)
}

## Recruit size pdf
c_0z1 <- function(z1, m.par)
{
    mu <- m.par["rcsz.int"]
    sig <- m.par["rcsz.sd"]
    p.deRecr <- dnorm(z1, mean = mu, sd = sig)              # pdf of a size z1 recruit
    return(p.deRecr)
}
```

```{r, echo=FALSE, results=FALSE}
# IBM has no density-dependence --> expect population to grow/decline exponentially - let's test this 
#**what is the difference between the IBM and IPM??**

# calculate iteration matrices 
IPM.true <- mk_K (nBigMatrix, m.par.true, -2.65, 4.5) #(# meshpts, parameter vector, lower integration limit, upper integration limit)
IPM.est <-mk_K (nBigMatrix, m.par.est, -2.65, 4.5)

# calculate real part of dominant eigenvalues (describe asymptotic rate of population increase)
Re (eigen (IPM.true$K)$values[1]) # eigen(calculates all eigenvalues ane eigenvectors of matrix and stores them in decreasing order)
Re (eigen (IPM.est$K)$values[1])

# for IBM, estimate finite rate of increase with lm pop size over t (but this is the IPM??)
fit.pop.growth <- lm (log(pop.size.t) ~ c(1:yr))
exp (coef(fit.pop.growth)[2])
c(1:yr)

# simplest condition for IPM to meet: some iterate of the kernel is a strictly positive function (check by making sure some multiplication of the iteration matric all entries are >0)
```

```{r, echo=FALSE, results=FALSE}
# code for generating figures in book
dev.new(); 
set_graph_pars("panel4")

## 1 - plot population density versus time...

plot(1:yr, log(pop.size.t[1:yr]), type = "l", xlab = "Time", ylab = "Population size")
mod.pop.growth <- lm(log(pop.size.t) ~ seq.int(1, yr))
abline(mod.pop.growth, col = "blue")

add_panel_label("a")

## ...roughly linear for log(Nt) vs time so exponential growth

## 2 - plot mean size versus time...
plot(1:yr, mean.z.t[1:yr], type = "l", xlab = "Time", ylab = "Mean plant size")
abline(h = mean.z.true, col = "red")
add_panel_label("b")

## ...mean size seems to settle down after a while

## 3 - plot mean flowering size versus time...
plot(1:yr, mean.fl.z.t[1:yr], type = "l", xlab = "Time", ylab = "Mean flowering plant size")
abline(h = mean.flowering.z.true, col = "red")
add_panel_label("c")

## ...mean flowering size seems to settle down after a while

## 4 - plot of density estimates at time 50 and the end
plot(density(sim.data$z), main = "", ylim = c(0, 0.4), xlab = "Plant size")
# lines(density(sim.data$z))
lines(IPM.est$meshpts, stable.z.dist.est/diff(IPM.est$meshpts)[1], col = "red")
add_panel_label("d")
```

```{r, echo=FALSE, results=FALSE}
# mean plant size and mean flowering plant size settle on some value - we can calc this value with the IPM

# MEAN PLANT SIZE
# calculate dominant eigenvector for iteration matrix (dominant right eigenvector is the stable distribution of the individual-level state variable (here is sizes, but could be age, stages, etc.))
meshpts <- IPM.true$meshpts # store defined meshpts for neater code
w.est <- Re(eigen(IPM.est$K)$vectors[, 1]) #find dominant eigenvector (w) of iteration matrix
stable.z.dist.est <- w.est/sum(w.est) # scale w so it becomes a discrete probability density
# proportion of individuals in each of the size categories centered on meshpts 
mean.z.est <- sum(stable.z.dist.est * meshpts) # calculate mean size of each size category (each meshpt * fraction of individuals whose size is that meshpt and taking the sum)
mean.z.est

# MEAN FLOWERING PLANT SIZE
wb.est <- p_bz(meshpts, m.par.est) * w.est
stable.flowering.dist.est <- wb.est / sum(wb.est)
mean.flowering.z.est <- sum(stable.flowering.dist.est * meshpts)
mean.flowering.z.est
```
##################################################################################
END OF SKIP







# 3.3 Case study 1B: Monocarp life cycle properties and events

Use the Oenothera IBM, but set the probability of recruitment ($p_r$) so that $\lambda=1$
- $\lambda=1$ means we don't need to discount by population size 
    * not exactly sure what this means, says like in Soay sheep example, revisit ****this***

# read in
1) Monocarp Demog Funs.R <- demographic function that describes how size influences rate; functions that describe how size influences rate
2) Monocarp Simulate IBM.R <- simulates IBM
3) Monocarp Events Calculations.R <- implements various event calculations

With monocarps, reproduction is fatal and there are 2 ways to die.

We want to test the various calculations implemented in Monocarp Events Calculations.R <- Use the true parameter values to define the IPM
We want to calculate various metrics stratified by age <- Collect a large sample of individual fates

We will need survival and reproduction iteration matrices
- already constructed IPM using mk_K(nBigMatrix, m.par.true, L, U)
- store iteration matrices in P, F
- define $\texttt{e <- matrix(1, nrow=1, ncol=nBigMatrix)}$ <- use this for summing down the columns of matrices
- define $\texttt{offspring.prob <- h * c_0z1(meshpts, m.par)}$ <- probability distributionof offspring sizes, which we will use as the initial cohort $c(z)$

## 3.3.1 Population Growth
$\lambda$ deals with changes in population size from 1 time step to the next
$R_0$ deals with changes in population size 1 generation to the next
Both can be calculated as eigenvalues of the appropriate kernel 






The above doesn't work for me, so I found this solution instead (used specific filepaths so please commented out or edit to run on your computer!)
## Nevermind, this broke too
I was able to run all of this code, then trying to knit everything broke regarding my RProject
```{r setup alternative}
# ## Make a function that takes a path and another function
# ## and returns that same function with the path pre-progammed in
 pathit <- function(FUN, path){
     function(file, ...){
         FUN(file=file.path(path, file), ...)
     }
 }
# 
# ## generate new functions that have the path pre-programmed in
 mono.sim.IBM <- pathit(source, "~/Dropbox/dissertation research/IPM/ipm_book/Rcode/c2")
 mono.dem.fns <- pathit(source, "~/Dropbox/dissertation research/IPM/ipm_book/Rcode/c2")
# 
source("Standard Graphical Pars.R")
mono.sim.IBM("Monocarp Simulate IBM.R")
mono.dem.fns("Monocarp Demog Funs.R")

```

## Generate artificial data by simulating IBM from Chapter 2 
## everything broke
```{r}
rm(list = ls())

library(doBy)
library(parallel)
set.seed(53241986)

source("~/Dropbox/dissertation research/IPM/ipm_book/Rcode/utilities/Standard Graphical Pars.R")
init.pop.size <- 100000
n.yrs <-50

m.par.true["p.r"] <- 0.005325442  #changed p.r so R0=1
#source("../c2/Monocarp Simulate IBM.R")
source("~/Dropbox/dissertation research/IPM/ipm_book/Rcode/c2/Monocarp Simulate IBM.R")


cat(pop.size.t,"\n") 
#source("~/Dropbox/dissertation research/IPM/ipm_book/Rcode/c2/Monocarp Simulate IBM.R")
source("~/Dropbox/dissertation research/IPM/ipm_book/Rcode/c2/Monocarp Demog Funs.R")

# no idea why I have to include this twice but R won't run it otherwise
m.par.true["p.r"] <- 0.005325442  #changed p.r so R0=1
#source("../c2/Monocarp Simulate IBM.R")
source("~/Dropbox/dissertation research/IPM/ipm_book/Rcode/c2/Monocarp Simulate IBM.R")


## trim an initial transient off the simulation
sim.data <- sim.data[sim.data$yr > 10,];
sim.data$yr <- sim.data$yr-10;
```

## Section 1 - Construct IPM kernels; assumes the IBM has been run and results stored in sim.data 
```{r}



nBigMatrix <- 250; 

min.size <- with(sim.data, min(z))
max.size <- with(sim.data, max(z))

# construct iteration matrix
IPM.true <- mk_K(nBigMatrix, m.par.true, -3.5, 5.5)

meshpts <- IPM.true$meshpts
h <- diff(meshpts)[1]
```

Calculate $\lambda$
```{r lambda}
# get the eigenvalues
lambda0 <- Re(eigen(IPM.true$K,only.values = TRUE)$values[1]) #only.values = TRUE specifies to only calculate eigenvalues (not eigenvectors)

fit.pop.growth <- lm(log(pop.size.t[-(1:10)])~c(11:yr))
lambda.hat <- exp(coef(fit.pop.growth)[2])
cat("Growth rate: IPM ",lambda0,"  sim ",lambda.hat, "\n")

# plot this
```

## IPM kernels for the calculations 
```{r}
# define the iteration matricies
P <- IPM.true$P # survival iteration matrix
F <- IPM.true$F # fecundityiteration matrix
```

kernel $R=FN$ 
```{r}
# calculate fundamental matrix (fundamental operator )
N <- solve(diag(nBigMatrix)-P)
# diag(nBigMatrix) creates ID matrix (I)
# solve inverts (I-P) matrix


R <- F %*% N # R = kernel -> projects the population from 1 generation to the next and 
```

Compute $R0$
```{r}
R0 <- abs(eigen(R)$values[1]) # calc dominant eigenvalue of R
cat("R0=",R0,"\n")
```

In IPM, offspring size is independent of parental size - can calculate $R_0$ by averaging per-capita fecundity with respect to size at birth
```{r}
# Define a probability distribution for offspring sizes
offspring.prob <- h*c_0z1(meshpts,m.par.true); 
sum(offspring.prob); # should be 1, and it is. 

#Define the e vector which we will use - calc sum of each column (250 columns)
e <- matrix(1,nrow=1,ncol=dim(P)[1]) # P=survival iteration matrix
# returns a vector 
```

```{r}
# With mixing at birth, we can calculate R0 directly as the 
# mean per-capita lifetime fecundity across a birth cohort
R0 <- sum((e %*% R) * offspring.prob)  
cat("R0=",R0,"\n");
```
This is integrating nBigMatrix=251 **what does this mean**

# 3.3.2 Mortality: age and size at death calculations

Calculate $l_a$ (probability of survival to age $a$):

$\tilde{f_a}=\frac{1}{\tilde{l_a}}\int_z \int_z FP^a (z',z)c(z)\ \mathrm{d}z\ \mathrm{d}z'$
$c(z)$ = initial state distribution (probability distribution)
Can calculate the average survival and per capita fecundity for a cohort of individuals with any $c(z)$:
${\tilde{l_a}}=\langle l_{a},c \rangle=\int_z l_{a}(z)c(z)\ \mathrm{d}z$ 
average value with respect to probability distribution, c: $\langle l_{a},c \rangle = \langle eP^a,c \rangle$ 

```{r}
sum ((e %*% P) * offspring.prob) # probability of surviving to age 1
```

Remember that $eP$ is $eP=\int_z eP(z',z) \mathrm{d}z'$ which is 1 iteration.
We already multiplied $P$ by $h$ so now, we integrate over $z'$ at each mesh point $z_i$ in 1 calculation.
```{r}
#Lots of the calculations involve multiplying e by P let's see what this does
cat(round((e %*% P),4)[1:10],"\n") 
```
$eP$ is a function of $z$: above code returns vector, with each entry containing $\int_z eP (z',z_{i}) \mathrm{d}z'$

SUmming down columns of survival iteration matrix $P$ corresponds to summing over all sizes (states) an individual can move to next year. Because all individuals that survive will also have a size next year, $\int_z eP(z',z) \mathrm{d}z'$ = probability of survival.

An alternative way to do the above is to calculate the probability of being alive next year for each of the meshpts (remembering flowering is fatal):
```{r}
cat(round(s_z(meshpts,m.par.true)*(1-p_bz(meshpts,m.par.true)),4)[1:10],"\n")
#does the same.
```

For a cohort with size distribution $c(z)$, the average of $eP$
**I have no idea what they are saying in this next sentence**

## Age-specific vital rates: 

alive means you didn't flower or die (Repr==0 & Surv==1), so this gives the probability of survival at each age
```{r}
summaryBy(alive~age,data=sim.data)

# #How many individuals in each age class?
summaryBy(alive~age,FUN=length,data=sim.data)
# #So we have plenty up to about age 5

# create empty vector to hold survivorship curve 
# la is the probability of surviving to age a
la <- rep(NA,20)

# calculate la[1], survival of age 0 individuals to age 1
la[1] <- sum((e %*% P)*offspring.prob)
la
```

For other age classes, we need to calculate $P_{a}$ = $a^{th}$ power of matrix $P$:
```{r}
#Later survivorships require P^a so let's do the calculation recursively
Pa <- P
for(a in 2:20){
	Pa <- Pa %*% P
	la[a] <- sum((e %*% Pa) * offspring.prob)
}

la <- c(1,la)
```

 to calculate the probability of survival from age a to age a+1 
```{r}
# we calculate la[a+1]/la[a]
pa <- la[2:21]/la[1:20]

# which compares well with the simulation data (plotted below) 
sim.pa <- summaryBy(alive~age,data=sim.data,keep.names=TRUE)
```

 Calculate the age-specific expected fecundity, fa
```{r}
Pa <- P
fa=rep(NA,20)
fa.0=sum((e %*% F)*offspring.prob)
fa[1]=sum((e %*% F %*% P)*offspring.prob) # calculate for age 1 individuals

for(a in 2:20){
	Pa=Pa %*% P
	fa[a]= sum((e %*% F %*% Pa)*offspring.prob)
}

# calculate survival probability
fa <- c(fa.0,fa)

fa <- fa/la
round(fa,4)

# Compute age-specific fecundity in simulation 
sim.data$Recruits = m.par.true["p.r"]*sim.data$Seeds
sim.data$Recruits[is.na(sim.data$Seeds)] <- 0
sim.fa <- summaryBy(Recruits~age,data=sim.data,keep.names=TRUE)
sim.fa
#summaryBy(Repr~age,FUN=sum,data=sim.data)

### Compare theoretical pa and fa with simulation results 
#(height=4,width=8); set_graph_pars("panel2"); 
par(mfrow=c(1,2))
plot(alive ~ age, data = sim.pa[1:10,],ylim=c(0,1),pch=19,
	xlab=expression("Age"),
	ylab=expression("p"[a]))
lines(0:9,pa[1:10])
add_panel_label("a")

plot(Recruits ~ age, data = sim.fa[1:10,],pch=19,
	xlab=expression("Age"), 
	ylab=expression("f"[a]))

lines(0:9,fa[1:10])
add_panel_label("b")
```

Calculate generation time using definitions from Section 3.1.3
```{r}
T1 <- log(R0)/log(lambda0)
T2 <- sum((1:21)*fa*la)/sum(fa*la)
cat("Generation time log(R0)/log(lambda) = ",T1,"\n")
cat("Generation time from age-specific rates = ",T2,"\n")

w <- Re(abs(eigen(IPM.true$K)$vectors[,1])) # get dominant eigenvector

off.dist <- F %*% w / (sum(F %*% w))
```
Differences between the 2 measures of generation time are small!

### Age at death - Mean and variance in lifespan

For average lifespan, need to evaluate $\tilde{\eta}=\langle eN,c \rangle$:
```{r}
# For the offspring distribution calculate the expected lifespan
# To get the mean age at death we subtract 1 from mean lifespan
# because by our convention lifespan is the number of censuses at which an
# individual is alive, so if you die with lifespan 1 your age at death is 0.    
mean.age.death <- round(sum((e %*% N)*offspring.prob) -1,4) 

# sum down fundamental matrix (N) to give expected lifespan at each meshpt, then take average with respect to probability distribution of offspring sizes

#check with the simulation, by calculating the mean age at death
mean.age.death.sim <- round(with(sim.data,mean(age[alive==0])),4)

# calculate variance in lifespan
Var.nu <- round(sum((e %*% (2 * N %*% N - N))* offspring.prob)  - sum((e %*% N * offspring.prob ))^2,4)

# check with the simulation - adding a constant doesn't change the variance so we don't need to
# subtract 1
Var.nu.sim <- round(with(sim.data,var(age[alive==0])),4)

cat("Age at death","\n")
cat("Theory: mean=",mean.age.death, " variance=",Var.nu,"\n"); 
cat("Simulation: mean=",mean.age.death.sim, " variance=",Var.nu.sim,"\n");
```

plot simulation vs theory
```{r}
ymax <- max(sim.data$yr);
Vnu.sim <- numeric(ymax)
for(j in 1:ymax) {
  Vnu.sim[j] <- with(sim.data,var(age[(alive == 0)& (yr==j)]))
}

par(mfrow=c(1,1),mar=c(5,4,3,1),mgp=c(2.5,1,0));
plot(2:ymax, Vnu.sim[-1],xlab="Length of simulation",ylab="Var(age at death)",
     ylim=c(0,1.1*Var.nu),bty="l",pch=1);
abline(h=Var.nu,lty=2);
title(main="Monocarp Var(lifespan): sim vs. theory");

py <- Vnu.sim[10:ymax];
px <- 1/(10:ymax);
fitVnu <- lm(py~px);
points(10:ymax,fitVnu$fitted,type="l",col="red");
cat(Var.nu,fitVnu$coef[1],"\n");

legend("bottomright",legend=c("Simulation","Theory"),
       lty=c(1,2),pch=c(1,NA),bty="n")
```

### Size at death - Mean and variance in size

**insert equation for omega**

Probability of dying - remember there are 2 ways (flowering or natural):
```{r}
# multiply each column in fundamental matrix N by vector contaiming probability of death at each meshpt
omega <-  (meshpts*(p_bz(meshpts,m.par.true) 
          + (1-p_bz(meshpts,m.par.true))*(1-s_z(meshpts,m.par.true)))) %*% N

#  how can you die? You flower with probability p_bz, and if you don't 
#  flower (1-p_bz) you die with probability (1-s_z)
mean.size.death <- sum(omega * offspring.prob)
round(mean.size.death,4)
with(sim.data,mean(z[alive==0])) 
```

Calculate the size at death kernel
```{r}
Omega <- (p_bz(meshpts,m.par.true) + (1-p_bz(meshpts,m.par.true))*(1-s_z(meshpts,m.par.true))) * N

# each of the columns defines a probability distribution and so should sum to 1, let's check
round(e %*% Omega,4)
#all 1's as it should be.
```

Calculate distribution of size at death
```{r}
# then the distribution of sizes at death for the offspring distribution is
dist.size.death <- (Omega %*% offspring.prob)

# again this should be a probability distribution and so sum to 1, let's check
round(e %*% dist.size.death,4)
# so it's all good, let's calculate some moments

# 1st the mean is
mean.size.death <-round(sum(dist.size.death*meshpts),4)
mean.size.death.sim <- round(with(sim.data,mean(z[alive==0])),4)  

# 2nd the variance is
Var.size.death <- round(sum(dist.size.death*meshpts*meshpts) - 
          sum(dist.size.death*meshpts)*sum(dist.size.death*meshpts),4)
          
Var.size.death.sim <- round(with(sim.data,var(z[alive==0])),4)

# compare simulation to theory
cat("Size at death","\n")
cat("Theory: mean=",mean.size.death, " variance=",Var.size.death,"\n")
cat("Simulation: mean=",mean.size.death.sim, " variance=",Var.size.death.sim,"\n")
# very close to each other!
```

**************
left off at 3.3.3
***************
