---
title: "R Notebook version of Monocarp Calculations.R"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

## Use the Monocarp IBM to illustrate the construction of an IPM

```{r}
rm(list = ls(all = TRUE))

library(doBy)
set.seed(53241986)

# Set simulation parameters
init.pop.size <- 250
n.yrs <- 100

#source(" /Standard Graphical Pars.R")

source("/Users/BartMBP/GitHub/first-edition/Rcode/c2/Monocarp Demog Funs.R")
source("/Users/BartMBP/GitHub/first-edition/Rcode/c2/Monocarp Simulate IBM.R")
cat(pop.size.t, "\n")
```
## trim an initial transient off the simulation
```{r}
sim.data <- sim.data[sim.data$yr > 10, ]
sim.data$yr <- sim.data$yr - 10
head(sim.data)
```

# Extract 1000 observations to use as our data set
```{r}
sample.index <- sample(1:nrow(sim.data), size = 1000, replace = FALSE)
sim.data <- sim.data[sample.index, ]
dim(sim.data)
```

## ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~## Section 1 - Fit statistical models to simulated data
## ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

Let's use sim data to construct an IPM, assuming we know the order of events first a bit of house keeping for plotting - chop up the sizes into classes and sort the data by size
```{r}
sim.data <- transform(sim.data, z.classes = cut(z, 20))
sim.data <- sim.data[order(sim.data$z), ]
head(sim.data)
```

## fit the functions and plot some graphs

## 1 - growth
```{r}
mod.Grow <- lm(z1 ~ z, data = sim.data)
summary(mod.Grow)

```

```{r}
plot(z1 ~ z, data = sim.data, pch = 16, cex = 0.25, 
     xlab = expression("Size t, " * italic(z)), 
     ylab = expression("Size t+1, " * italic(z) * "'"), 
     main="Growth")
abline(mod.Grow, col = "red")
lines(z,z,col="gray")
```


## 2 - flowering

```{r}
mod.Repr <- glm(Repr ~ z, family = binomial, data = sim.data)
summary(mod.Repr)
```


```{r}
Repr.ps <- summaryBy(z + Repr ~ z.classes, data = sim.data)
plot(Repr ~ z, data = sim.data, cex = 0.7, 
     xlab = expression("Size t, " *italic(z)), 
     ylab = "Probability of flowering", col="gray")
points(Repr.mean ~ z.mean, pch = 19, col="black", data = Repr.ps)
lines(fitted(mod.Repr) ~ z, data = sim.data, col = "red")
```


## 3 - survival
```{r}
sim.data.noRepr <- subset(sim.data, Repr == 0)
mod.Surv <- glm(Surv ~ z, family = binomial, data = sim.data.noRepr)
summary(mod.Surv)
```

```{r}
surv.ps <- summaryBy(z + Surv ~ z.classes, data = sim.data.noRepr, na.rm = TRUE)

plot(Surv ~ z, data = sim.data.noRepr, col="gray", cex = 0.7,
     xlab = expression("Size t, " *italic(z)), 
     ylab = "Probability of survival")
points(Surv.mean ~ z.mean, data = surv.ps, pch = 19)

lines(fitted(mod.Surv) ~ z, data = sim.data.noRepr, col = "red")

```


## 4 - seed production
```{r}
sim.data.Repr <- subset(sim.data, Repr == 1)
mod.Seeds <- glm(Seeds ~ z, family = poisson, data = sim.data.Repr)
summary(mod.Seeds)
```

```{r}
plot(log(Seeds) ~ z, data = sim.data.Repr, pch = 19, 
     xlab = expression("Size t, " *italic(z)), 
     ylab = "Seeds production")

abline(mod.Seeds, col = "red")

```



## 5 - recruit size
```{r}
sim.data.Rec <- subset(sim.data, age == 0)

mod.Rcsz <- lm(z ~ 1, sim.data.Rec)
summary(mod.Rcsz)
```

```{r}
hist(sim.data.Rec$z, xlab="Size of Recruits")
```



## 6 - establishment probability 
```{r}
p.r.est <- as.numeric(Recr)/sum(Seeds, na.rm = TRUE)
p.r.est
```


## Finally, store the estimated parameters
```{r}
m.par.est <- c(surv = coef(mod.Surv), flow = coef(mod.Repr), grow = coef(mod.Grow), 
               grow.sd = summary(mod.Grow)$sigma, rcsz = coef(mod.Rcsz), rcsz.sd = summary(mod.Rcsz)$sigma, seed = coef(mod.Seeds), p.r = p.r.est)

m.par.est
```


This is just giving the parameters the same names as the true parameters
```{r}
names(m.par.est) <- names(m.par.true)
m.par.est
```



## Section 2 - Construct Kernels and projection population size
this is just finding the min and max of z in sim.data
```{r}
nBigMatrix <- 250
min.size <- with(sim.data, min(z))
min.size
max.size <- with(sim.data, max(z))
max.size
```
so let's set the lower and upper limits for the size range at -2.65
, and 4.5 so slightly smaller/bigger than observed.


#From Monocarp Demog Funs.R

## Define the various demographic functions.
We pass a vector of parameters "m.par" to each function. This makes it easier to compare the results of different paramter sets, say the true values and estimated ones


## Growth function, given you are size z now returns the pdf of size z1 next time
```{r}
G_z1z <- function(z1, z, m.par)
{
    mu <- m.par["grow.int"] + m.par["grow.z"] * z           # mean size next year
    sig <- m.par["grow.sd"]                                    # sd about mean
    p.den.grow <- dnorm(z1, mean = mu, sd = sig)             # pdf that you are size z1 given you were size z
    return(p.den.grow)
}

```


## Survival function, logistic regression
```{r}
s_z <- function(z, m.par)
{
    linear.p <- m.par["surv.int"] + m.par["surv.z"] * z  # linear predictor
    p <- 1/(1+exp(-linear.p))                            # logistic transformation to probability
    return(p)
}

```


## Probability of flowering function, logistic regression
```{r}

p_bz <- function(z, m.par)
{
    linear.p <- m.par["flow.int"] + m.par["flow.z"] * z      # linear predictor
    p <- 1/(1+exp(-linear.p))                                # logistic transformation to probability
    return(p)
}

```

## Seed production function
```{r}
b_z <- function(z, m.par)
{
    N <- exp(m.par["seed.int"] + m.par["seed.z"] * z)    # seed production of a size z plant
    return(N)
}
```


## Recruit size pdf
```{r}
c_0z1 <- function(z1, m.par)
{
    mu <- m.par["rcsz.int"]
    sig <- m.par["rcsz.sd"]
    p.deRecr <- dnorm(z1, mean = mu, sd = sig)              # pdf of a size z1 recruit
    return(p.deRecr)
}
```



# Section 2 - Functions to build IPM kernels P, F, and K

## Define the survival kernel
```{r}
P_z1z <- function (z1, z, m.par) {

    return((1 - p_bz(z, m.par)) * s_z(z, m.par) * G_z1z(z1, z, m.par))

}
```


## Define the fecundity kernel
```{r}
F_z1z <- function (z1, z, m.par) {

    return( p_bz(z, m.par) * b_z(z, m.par) * m.par["p.r"] * c_0z1(z1, m.par))

}
```

## Make Kernal
```{r}
mk_K <- function(m, m.par, L, U) {

	# mesh points 
	h <- (U - L)/m
	meshpts <- L + ((1:m) - 1/2) * h
	P <- h * (outer(meshpts, meshpts, P_z1z, m.par = m.par))
	F <- h * (outer(meshpts, meshpts, F_z1z, m.par = m.par))
	K <- P + F
	return(list(K = K, meshpts = meshpts, P = P, F = F))
}

```


##An aside:  what does 'outer' do? 

'outer' takes two vectors and a function (that itself takes two arguments) and builds a matrix by calling the given function for each combination of the elements in the two vectors.

```{r}
x <- 0:2; # three columns
y <- 0:4; # five rows

m <- outer (
   y,     # First dimension:  the columns (y)
   x,     # Second dimension: the rows    (x)
   function (x, y)   x+2*y
);

m

```




```{r}
IPM.true <- mk_K(nBigMatrix, m.par.true, -2.65, 4.5)
IPM.est <- mk_K(nBigMatrix, m.par.est, -2.65, 4.5)

Re(eigen(IPM.true$K)$values[1])

Re(eigen(IPM.est$K)$values[1])

```


```{r}
fit.pop.growth <- lm(log(pop.size.t) ~ c(1:yr))

exp(coef(fit.pop.growth)[2])

```

```{r}
meshpts <- IPM.true$meshpts

w.est <- Re(eigen(IPM.est$K)$vectors[, 1])
stable.z.dist.est <- w.est/sum(w.est)
mean.z.est <- sum(stable.z.dist.est * meshpts)
mean.z.est

```

```{r}
w.true <- Re(eigen(IPM.true$K)$vectors[, 1])
stable.z.dist.true <- w.true/sum(w.true)
mean.z.true <- sum(stable.z.dist.true * meshpts)
mean.z.true

```

```{r}
wb.est <- p_bz(meshpts, m.par.est) * w.est
stable.flowering.dist.est <- wb.est/sum(wb.est)
mean.flowering.z.est <- sum(stable.flowering.dist.est * meshpts)
mean.flowering.z.est

```

```{r}
wb.true <- p_bz(meshpts, m.par.true) * w.true
stable.flowering.dist.true <- wb.true/sum(wb.true)
mean.flowering.z.true <- sum(stable.flowering.dist.true * meshpts)
mean.flowering.z.true

```


## 1 - plot population density versus time...
```{r}
plot(1:yr, log(pop.size.t[1:yr]), type = "l", xlab = "Time", ylab = "Population size")
mod.pop.growth <- lm(log(pop.size.t) ~ seq.int(1, yr))
abline(mod.pop.growth, col = "blue")
```



## ...roughly linear for log(Nt) vs time so exponential growth

## 2 - plot mean size versus time...
```{r}
plot(1:yr, mean.z.t[1:yr], type = "l", xlab = "Time", ylab = "Mean plant size")
abline(h = mean.z.true, col = "red")

```

## ...mean size seems to settle down after a while

## 3 - plot mean flowering size versus time...
```{r}
 plot(1:yr, mean.fl.z.t[1:yr], type = "l", xlab = "Time", ylab = "Mean flowering plant size")
abline(h = mean.flowering.z.true, col = "red")
```
...mean flowering size seems to settle down after a while

## 4 - plot of density estimates at time 50 and the end
```{r}
plot(density(sim.data$z), main = "", ylim = c(0, 0.4), xlab = "Plant size")
# lines(density(sim.data$z))
lines(IPM.est$meshpts, stable.z.dist.est/diff(IPM.est$meshpts)[1], col = "red")

```


